### 2020-06-04 report/sections/server/node/efsm.tex
Report: progress on chapter 4 runtime

For the time being,
it is sufficient to understand that for each EFSM state,
in addition to the API defined under the
\texttt{Handler} namespace, it also has a wrapper API defined under
the \texttt{Implementation} namespace (\cref{lst:nodeefsmimplementation}),
which defines the \texttt{type}
discriminant property internally. This explains why the successor
state encodings in 
\cref{lst:addersvrsendhandler,lst:addersvrreceivehandler}
were defined as such.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
abstract class ISend { type: 'Send' = 'Send'; ... }
abstract class IReceive { type: 'Receive' = 'Receive'; ... }

export namespace Implementation {

	export class S51 extends IReceive {
		constructor(private handler: Handler.S51) { super(); }
		...
	}
	...
};
\end{lstlisting}
\captionof{lstlisting}{Discriminated Unions in EFSM 
for Server-Side Endpoints}
\label{lst:nodeefsmimplementation}
\end{figure}

>>>>

For the time being,
it is sufficient to understand that for each EFSM state,
in addition to the API defined under the
\texttt{Handler} namespace, it also has a wrapper API defined under
the \texttt{Implementation} namespace (\cref{lst:nodeefsmimplementation}),
which defines the \texttt{type}
discriminant property internally (via inheritance).
This explains why the successor
state encodings in 
\cref{lst:addersvrsendhandler,lst:addersvrreceivehandler}
were defined as such.
The type \texttt{Implementation.Type} is the discriminated union,
which is used by the runtime.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
abstract class ISend { type: 'Send' = 'Send'; ... }
abstract class IReceive { type: 'Receive' = 'Receive'; ... }
abstract class ITerminal { type: 'Terminal' = 'Terminal'; ... }

export namespace Implementation {
	export type Type = ISend | IReceive | ITerminal;

	export class S51 extends IReceive {
		constructor(private handler: Handler.S51) { super(); }
		...
	}
	...

};
\end{lstlisting}
\captionof{lstlisting}{Discriminated Unions in EFSM 
for Server-Side Endpoints}
\label{lst:nodeefsmimplementation}
\end{figure}

### 2020-06-04 report/sections/server/node/runtime.tex
Report: progress on chapter 4 runtime

We define the session runtime in Svr.ts,
named after the \trole{Svr} endpoint.
It exposes a \textbf{public API}
(\cref{lst:noderuntimepublicapi})
with seams for the developer to pass in the WebSocket server 
and application logic
(i.e. their handler implementations).
It is the responsibility of the developer to construct the
WebSocket server and set it up to listen for incoming connections.
Internally, it keeps a \textit{private API}
for executing the EFSM, when all participants have 
joined the session.
The role of the public API is to manage incoming connections and
wait for all participants to join the session 
(\cref{subsection:noderuntimepublic}), before
handing off to the private API to execute the EFSM
(\cref{subsection:noderuntimeprivate}).

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2,title=Svr.ts]
export class Svr {
	constructor(wss: WebSocket.Server,
							initialState: Implementation.S51) { ... }
	...
}
\end{lstlisting}
\captionof{lstlisting}{Public API of Session Runtime for Server Endpoint}
\label{lst:noderuntimepublicapi}
\end{figure}

@@

We revisit the concept of using discriminated unions to distinguish
the type of EFSM state. The discriminated union is implemented by the
\texttt{Implementation} API which wraps the developer's
implementation of the  \texttt{Handler} API.

It is sufficient for the runtime to only discriminate the type of state
rather than the specific state. 
we can define \textit{state-specific}
details within the discriminated union

>>>>

We define the session runtime for the \trole{Svr} endpoint
of the \tprotocol{Adder} protocol in Svr.ts,
named after the endpoint.
It exposes a \textbf{public API}
(\cref{lst:noderuntimepublicapi})
with seams for the developer to pass in the WebSocket server 
and application logic
(i.e. their handler implementations).
It is the responsibility of the developer to construct the
WebSocket server and set it up to listen for incoming connections.
Internally, it keeps a \textit{private API}
for executing the EFSM, when all participants have 
joined the session.

The role of the public API is to manage incoming connections and
wait for all participants to join the session 
(\cref{subsection:noderuntimepublic}), before
handing off to the private API to execute the EFSM
(\cref{subsection:noderuntimeprivate}).

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,title=Svr.ts]
export class Svr {
	constructor(wss: WebSocket.Server,
				initialState: Implementation.S51) { ... }
	...
}

// Not exported to developer
class Session {
	private wss: WebSocket.Server;
	private initialState: Implementation.S51;
	private roleToSocket: RoleToSocket;
	...
}
\end{lstlisting}
\captionof{lstlisting}{Public and Private API of Session Runtime for Server Endpoint}
\label{lst:noderuntimepublicapi}
\end{figure}

@@

The \texttt{Session} class executes the EFSM.
We define a transition function, \texttt{next()}, 
parameterised by the current state,
which invokes the handler defined by the develoepr
and performs the required channel actions for non-terminal states:

\begin{itemize}
\item 
For \textbf{send} states, the handler will
return the label and payload to be sent, along with the
successor state implementation. The transition function
should construct and send the message, 
and transition to the successor state.

\item
For \textbf{receive} states, we change the
message event listener on the WebSocket to
pass the incoming message to the handler.
The handler will return the successor state,
which the runtime can transition to.

\end{itemize}

We conceptualise this in \cref{lst:noderuntimesimple}.
The discriminated union lets the runtime figure out
the type of the current state.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
next(state: Implementation.Type) {
	switch (state.type) {
		case 'Send': {
			const [label, payload, succ]: (*@\hl{???}@*) = state.handler;	 (*@\label{line:noderuntimebadtype1}@*)
			this.send((*@\hl{???}@*), label, payload); (*@\label{line:noderuntimeunknownrole}@*)		
			return this.next(succ);
		}
		case 'Receive': {
			this.wss.onmessage = ({ data }) => {
				const { label, payload } = JSON.parse(data) as (*@\hl{???}@*); (*@\label{line:noderuntimebadtype2}@*)
				const succ: (*@\hl{???}@*) = state.handler[label](...payload); (*@\label{line:noderuntimebadtype3}@*)
				return this.next(succ);
			}
		}
		case 'Terminate': { return; }
	}
}
\end{lstlisting}
\captionof{lstlisting}{Conceptual EFSM Transition Function 
for Server-Side Endpoint}
\label{lst:noderuntimesimple}
\end{figure}

However, we still face problems with resolving types, as highlighted.
Just because we know that the current state is a send state,
we do not know \textit{which} particular state it is,
so we cannot accurately type the handler (\cref{line:noderuntimebadtype1}).
The same problem is amplified for the receive state:
we need to know the specific receive state in order
to correctly serialise the message (\cref{line:noderuntimebadtype2})
and interpret the successor state (\cref{line:noderuntimebadtype3}).
We see another problem with handling send states:
because we do not know the specific send state, we do not know which
participant to send the message to (\cref{line:noderuntimeunknownrole}).

These all reduce to the same core problem: the runtime
needs to know the specific state at compile-time\footnote{
Whether TypeScript ``compiles'' or ``transpiles''
(or even \textit{``transcompiles''} \cite{transcompiles}) to JavaScript
is not relevant to our work; we stick with compilation
and keep our terminology consistent.}.
We solve this through \textit{runtime polymorphism} instead, since
the specific type of \texttt{state} is known at runtime.
For each type of state, we define a common API that can be invoked
by the EFSM transition function. To achieve runtime polymorphism,
each concrete state must provide a specific implementation: 
this motivates our design for defining the discriminated union
using abstract classes with abstract methods 
(\cref{lst:nodeabstractclass}).

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
abstract class ISend {
	type: 'Send' = 'Send';
	abstract performSend(
		next: EfsmTransitionHandler,
		sendMessage: (role: Roles.Peers, label: string, payload: any[])
			=> void,
	): void;
};

abstract class IReceive {
	type: 'Receive' = 'Receive';
	abstract prepareReceive(
		next: EfsmTransitionHandler,
		register: (from: Roles.Peers, messageHandler: MessageHandler)
			=> void,
	): void;
};
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:nodeabstractclass}
\end{figure}

Using this approach,
we can pass the transition function and channel actions 
from the \texttt{Session} runtime class
to the individual state \texttt{Implementation} classes:
these are also generated by \fancyname{NodeTS}, so we
guarantee linear usage of channel resources by construction as well.
This \textit{significantly} simplifies the design of the runtime
(\cref{lst:noderuntime}),
because the transition function no longer needs to know the type 
of the specific state at compile-time.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
next(state: Implementation.Type) {
	switch (state.type) {
		case 'Send':
			return state.performSend(this.next, this.send);
		case 'Receive':
			return state.prepareReceive(this.next, this.register);
		case 'Terminal':
			return;
	}
}
\end{lstlisting}
\captionof{lstlisting}{Final EFSM Transition Function for Server-Side
Endpoint}
\label{lst:noderuntime}
\end{figure}

By passing the specific functions defined in the runtime
to the individual EFSM states, we can visualise our runtime
as a \textit{message passing} abstraction 
(\cref{fig:noderuntimeefsm}): the runtime uses the common
\texttt{performSend()} or \texttt{prepareReceive()} API
to delegate to the specialised implementation, which will in turn
ask the runtime to perform specialised channel actions using
the parameterised methods, and finally delegate back to
the runtime to transition to the specific successor state.

We illustrate
